Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-18 paper/paper.tex
committing

A {\em liveness environment} is a mapping from variables to subsets of
$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
writing  $\{x.\acar\acdr,   x.\acdr,  y.\epsilon\}$  instead   of  $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  each
program  point $\pi$:  the liveness  at $\pi:e$  is the  liveness just
before executing $e$.
%%
%% Due to the inherent uncertainty in
%% determining the  point of evaluation  in lazy languages,  we associate
%% liveness environments with certain kinds  of paths in the program. All
%% programs points lying on the path share the same liveness environment.
%% We explain this point later.
%% 
A  notion  related  to  liveness  is {\em  demand}.   The  demand on
expression  $e$   is  again  an   access  path---that  subset   of  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.  So,  for example given  a demand $\sigma$ and  the expression
$\pi: (\RETURN\ x)$, the liveness at $\pi$ is exactly $x.\sigma$.  The
classical  analogy of  this  is  in {\em  strong  liveness}, where  an
assignment node $n: x:=y+z$ causes $y$  and $z$ to be live on entry to
$n$ if (and only if) $x$ is  live at exit of $n$---the liveness of $x$
at exit from $n$ becomes the demand on $y+z$.

@@

Note that liveness refers to  variables and fields, and not to $\CONS$
cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
future computation may refer to  the value $\ell$ of variable $x$, and
also to the $\CAR$ field of  location $\ell$.  In the absence of other
pointers to heap location $\ell$, we are certain that the $\CDR$ field
of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
would  be  garbage (again  provided  there  are  no other  aliases  to
$\ell$).

@@

To describe  the issues  in formulating liveness  of a  lazy language,
consider the function definition shown:

@@

To see  our way of  handling the second  issue, we trace  the liveness
analysis to  an extent. Assume  that the calls  to the $\pf$  place an
overall demand  of $\sigma$ on the  body of \pf.  on  the now describe
the process  of finding out the  liveness of each  variable defined in
the function  $\pf$ assuming  a symbolic demand  $\sigma$ on  it.  For
obvious reasons, this is also the  demand on the return values $u$ and
$y$, and thus  the liveness at the program  points $\pi_7$ and $\pi_8$
are $\lbrace  \pu.\sigma \rbrace$ and  $\{\py.\sigma\}$.  The liveness
at  $\pi_7$ and  $\pi_8$ are  propagated backwards  as  in traditional
liveness analysis.

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn generates  a liveness  of $\{\pw,~\pw.1\sigma\}$ at  $\pi_6$. The
liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
scope of  \pu. However, for simplicity  of exposition, we  do not kill
the variable in our  formulation.  This does not introduce imprecision
since  \pu\  is yet  to  come into  existence at  $\pi_6$ and  its
liveness will not be consulted by the garbage collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================
In an  eager  language, \pw\  would  not have  been live  after
 $\pi_7$, the point of its last use. In a lazy language, however, the
  definition  $\pu  \leftarrow  (\CDR~\pw)$,  does not  result  in  an
  evaluation  of  $(\CDR~\pw)$; instead  a  closure  is created.   The
  evaluation actually takes place while evaluating $(\RETURN~u)$. Thus
  it is necessary to record the liveness of $\pw$ at the program point
  $\pi_8$.
%% The second  difference is that  our \LET s  can be recursive  as in
%%   program point $\pi_1$.  In such a case, the liveness of a let has
%%   to be formulated as a fix-point computation.
  To address this issue, we  regard the body of the function as a
  tree and consider paths from the root of this tree to program points
  which     trigger     a     evaluation---\SIF\    conditions     and
  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
  analysis consists of two maps:  The map $\cal{M}$ maps a return path
  to a liveness  environment that is common to  all the program points
  in the path. And a second  map $\cal{P}$ maps a program point to the
  set of  return paths on  which the point  lies.  If a  program point
  happens  to fall  on  more than  one  such path,  then the  liveness
  environment of the  program point is the variable-wise  union of the
  liveness  environments  of the  individual  paths.   In the  example
  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.

  In  the  sequel we  shall  describe  how  to obtain  $\cal{M}$.  The
  formulation  of $\cal{P}$  is easy  and we  do not  describe  it any
  further.
  We observe that the ${ref}$ rules neither use ${\cal M}$ directly nor 
  create new labels. We only need to consider the rules for {\LET}, {\SIF}, and
  {\RETURN}.

>>>>

A {\em liveness environment} is a mapping from variables to subsets of
$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
writing  $\{x.\acar\acdr,   x.\acdr,  y.\epsilon\}$  instead   of  $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  each
program  point $\pi$:  the liveness  at $\pi:e$  is the  liveness just
before executing $e$.
%%
%% Due to the inherent uncertainty in
%% determining the  point of evaluation  in lazy languages,  we associate
%% liveness environments with certain kinds  of paths in the program. All
%% programs points lying on the path share the same liveness environment.
%% We explain this point later.
%% 
A  notion  related  to  liveness  is  {\em  demand}.   The  demand  on
expression  $e$   is  again  an   access  path---that  subset   of  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.   The  demand on  an  expression  generates  liveness for  its
constituent  variables.  So  for  example, given  a (non-null)  demand
$\sigma =  \lbrace \epsilon, \acar  \rbrace$ and the  expression $\pi:
(\CDR\  x)$, the liveness  at $\pi$  is $\lbrace  x.\epsilon, x.\acdr,
x.\acdr\acar \rbrace$.  This indicates  that to explore $\sigma$ parts
of the  result of $(\CDR\  x)$, one has  to first dereference  $x$ and
\CDR\  field of  the cons  cell referred  by $x$.  Note  that liveness
refers to  variables and  fields, and not  to $\CONS$ cells  (i.e.\ to
edges  in  the memory  graph,  not  to  locations themselves).   Hence
liveness of $\{x.\epsilon,x.\acar\}$ means that future computation may
refer to  the value  $\ell$ of  variable $x$, and  also to  the $\CAR$
field of location $\ell$.

%% The classical  analogy of this  is in
%% {\em strong liveness},  which can be expressed in  our terminology as:
%% $y$ and $z$  have a liveness of $\lbrace\epsilon\rbrace$  at the entry
%% to $n: x:=y+z$,  c$y$ and $z$ if  (and only if) $x$ has  a liveness of
%% $\lbrace\epsilon\rbrace$ at exit of  $n$---the liveness of $x$ at exit
%% from $n$ becomes the demand on $y+z$.

@@

%% Note that liveness refers to  variables and fields, and not to $\CONS$
%% cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
%% themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
%% future computation may refer to  the value $\ell$ of variable $x$, and
%% also to the $\CAR$ field of  location $\ell$.  In the absence of other
%% pointers to heap location $\ell$, we are certain that the $\CDR$ field
%% of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
%% arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
%% collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
%% would  be  garbage (again  provided  there  are  no other  aliases  to
%% $\ell$).

@@

@@

To see  our way of  handling the second  issue, consider the liveness
analysis of the following function:

Assume that the calls to the $\pf$ place an overall demand of $\sigma$
on the body  of \pf.  For obvious reasons, this is  also the demand on
the return  values $u$ and $y$,  and thus the liveness  at the program
points  $\pi_7$  and  $\pi_8$  are $\lbrace  \pu.\sigma  \rbrace$  and
$\{\py.\sigma\}$.  The liveness at  $\pi_7$ and $\pi_8$ are propagated
backwards as in traditional liveness analysis.

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn generates  a liveness  of $\{\pw,~\pw.1\sigma\}$ at  $\pi_6$. The
liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
scope of  \pu. However, for simplicity  of exposition, we  do not kill
  livenesses  of  variables  in  our formulation.   This  does  not
introduce any imprecision since \pu\  is yet to come into existence at
$\pi_6$  and  its  liveness  will  not be  consulted  by  the  garbage
collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================

In an  eager  language, \pw\  would  not have  been live  after
 $\pi_7$, the point of its last use. In a lazy language, however, the
  definition  $\pu  \leftarrow  (\CDR~\pw)$,  does not  result  in  an
  evaluation  of  $(\CDR~\pw)$; instead  a  closure  is created.   The
  evaluation actually takes place while evaluating $(\RETURN~\pu)$. Thus
  it is necessary to record the liveness of $\pw$ at the program point
  $\pi_8$, indeed at every program point  from $\pi_8$ up to the
  beginning of the program.
%% The second  difference is that  our \LET s  can be recursive  as in
%%   program point $\pi_1$.  In such a case, the liveness of a let has
%%   to be formulated as a fix-point computation.
  We thus  regard the body of the function as a
  tree and consider paths from the root of this tree to program points
  which     trigger     a     evaluation---\SIF\    conditions     and
  \RETURN\  expressions. We call  such paths  {\em e-paths}.   For the
  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path  to a  liveness environment  that is  common to  all  the program
points in the path. And a second map $\cal{P}$ maps a program point to
the set of  return paths on which the point lies.   If a program point
happens  to  fall  on more  than  one  such  path, then  the  liveness
environment of  the program  point is the  variable-wise union  of the
liveness  environments  of  the  individual  paths.   In  the  example
program, ${\cal M}(EP_1)=  \ldots$, ${\cal M}(EP_2)=\ldots$ and ${\cal
  M}(EP_3)= \ldots$.   Since ${\cal P}(pi_4)$ is  $\lbrace EP_1, EP_2,
EP_3 \rbrace$,  the liveness  of $pi_4$ is  \ldots.  In the  sequel we
shall describe how to  obtain $\cal{M}$.  The formulation of $\cal{P}$
is obvious  and we do  not describe it.   We observe that  the ${ref}$
rules neither use  ${\cal M}$ directly nor create  new labels. We only
need to consider the rules for {\LET}, {\SIF}, and {\RETURN}.
